\Large \bf Cyclone: Warp Speed Replication for Key Value Stores

### 2017-07-24 paper/paper.tex
More updates.

\subsection{Eliminating Data Movement}

Another key problem we had to solve with DPDK was teaching it about persistent
memory. DPDK allows one to plug in a memory allocator. We required the allocator
to support various features such as allowing multiple entities (the cpu core and
NIC driver) to hold reference to DPDK packets and allocate/free in parallel.

The NVM log is maintained as a circular log of fixed sized pointers (a pointer
log) to memory buffers - each of which holds a client request. This arrangement
is shown in Figure~\ref{fig:nvm_log} that also illustrates the advantage of such
an arragement. Adding a level of indirection in the NVM log allows us to
separate the FIFO ordered circular log being manipulated by the RAFT from
variable sized client being managed by the memory allocator of the packet
switching library we use. Another advantage is that it makes recovery from NVM
easy - appends to the circular pointer log are atomic and we can use the pointer
log to recover allocator state i.e. what pieces of NVM are currently in use by
the log.

@@

We now describe how we implement the RAFT consensus protocol to replicate the
NVM log across a set of replicas. The software packet switching idea we apply is
to separate the control plane from the data plane using the Data Plane
Development Kit (DPDK~\cite{dpdk}) library for direct userspace access to the
network interface.

\subsubsection{Dataplane Separation}
We use RAFT~\cite{raft} as the underlying protocol for log replication. RAFT
replicates a log across a set of replicas. Its operations are leader-based and
RAFT incorporates a leader election algorithm that elects a leader with the most
up to date log. The leader receives log entries from clients, appends them to the
persistent log (NVM log in this case) and multicasts the new entry to follower
replicas. On a response from enough follower replicas (constituting a majority
quorum) the leader replica can then execute the command in the log entry. In
our case this involves accessing the key value store and sending the response
back to the client.

We implement the common case flow in RAFT as a software packet switch using
DPDK. The dataplane now encompasses the Network Interface Card (NIC) and the CPU
caches with the NIC placing packet data directly in the CPU cache via
DDIO~\cite{ddio} - a standard feature of most NICs today. The dataplane for the
RAFT protocol is therefore separated from the control plane. The control plane
runs in a thread pinned to a dedicated CPU core and takes care of the protocol
state machine: checking for correct term, log index and that the replica
receiving the packet from the client is indeed the leader. On the follower side
it is responsible for checking that the leader is in the correct term before
accepting the packet by appending it to the NVM log and initiating a
response. In addition, the control plane is responsible for uncommon case
activities such as dealing with timeouts and leader elections.

The pseudocode in Figure~\ref{fig:control_plane} describes part of the control
plane in Cyclone organized as event handlers triggered on receiving a packet at
the leader. We focus on one event for illustration (and brevity): the event
where a request is received from a client. A key concern when we built Cyclone
was adhering as far as possible to the software packet switching principle of
separating the control plane from the dataplane. This means that the CPU should
avoid examining every byte of the packet as far as possible.

The sequence of operations when a packet is received from the client needs
careful design for efficient software packet switching. To illustrate how this
is achieved, Figure~\ref{fig:packet_layout} shows how Cyclone manipulates packet
layouts across the two steps of prepending a RAFT header and transmitting to
follower replicas. DPDK describes packets using an ``mbuf'' data
structure. Roughly speaking, an ``mbuf'' consists of a flat array of bytes
actually containing the packet and a fixed size piece of external metadata that
describes various aspects of the packet, most crucially a pointer to the start
and end of the packet in the byte array. DPDK's userspace drivers receive
packets from the NIC such that they are offset in the byte array by a
configurable amount referred to as ``headroom''. We strip off the existing
network headers in the packet and prepend RAFT related information specific to
each log entry in the headroom by shifting the start pointer appropriately.
These operations are standard enough for software packet switches that DPDK
provides convenient library calls for it. For the final step, we need to prepare
the packet for transmission to the various follower replicas. To do this we
prepare a different packet containing an ethernet header for \emph{each}
targeted replica and ``chain'' the data packet to each of these headers. Each
header is then separately handed off to the driver for transmission via the NIC,
carrying the data packet with it by association.

\begin{figure}
\begin{verbatim}
    event_receive_client_req()
    {
      if(!check_is_leader()) {
        drop_msg
        return
      }
      Prepend raft log term and index
      Persist to local log
      Transmit to follower replicas
    }
\end{verbatim}
\caption{Cyclone control plane fragment}
\label{fig:control_plane}
\end{figure}

@@

We note that by discarding stream oriented protocols such as TCP, the network
layer is no longer reliable or ordered. This is not a problem for RAFT that is
designed to tolerate network failures resulting in drops or reordering of
network packets. However, this rules out consensus protocols such as Zookeeper
Atomic Broadcast (ZAB~\cite{zab}) that depend on a reliable stream-oriented
network layer to function correctly.

Finally, we turn our attention to the persistence step in
Figure~\ref{fig:control_plane}. RAFT requires that the log entry be persisted
before it is multicast out to follower replicas. Since the NVM is directly
attached we do this by executing a cacheline flush ({\tt clflush}) instruction
for every cacheline in the packet and the pointer in the pointer buffer to
persist these via the memory bus. This is the one exception to our otherwise
clean separation between the control and data plane - necessitated by the fact
that at the moment NICs do not support persistent memory. This is not too
onerous a burden because we can use the newly introduced {\tt
  clflush-opt}~\cite{clflush_opt} instruction specifically intended to
efficiently flush to persistent memory without the overhead of the serialization
normally introduced by {\tt clflush}. This allows us to hit full memory
bandwidth on present generation platforms, a quantity in excess of 200 Gb/s per
core, well above the near term speeds of network interface cards.

@@

\subsubsection{Batching}
Although the control plane code only runs operations related to the RAFT state
machine it is still slow enough relative to the dataplane to add significant
overhead for each packet. We tackle this problem by applying another classic
software packet switching technique: batching - illustrated in
Figure~\ref{fig:batching}. We use a burst receive call available in the DPDK
userspace driver to receive a burst of client packets at a time. We then chain
these packets together and treat them as a single log entry from the perspective
of RAFT, amortizing the control plane overheads over the packets (at most 32 at
a time due to current driver limitations).

We note that batching in Cyclone does not involve a latency-throughput
tradeoff like in many other systems~\cite{ix-dataplane}. The batch receive call
we use in DPDK returns immediately with whatever number of packets is available,
including zero. We always flush the transmit buffer after every call to DPDK to
transmit packets to replicas. Therefore, we never tradeoff latency for
throughput when batching.

\subsection{Flash Log}

The fact that we can immedately absorb updates into an NVM log and only move
them to the flashlog in suitable units for block IO sidesteps a common problem
with logging to secondary storage - the need to do group commit. Logging systems
often wait to collect a batch of entries to write to secondary storage trading
latency for throughput. This is unnecessary in Cyclone as the use of directly
attached NVM with a two level log removes the need to make this tradeoff.

>>>>

We direct DPDK to use pages backed by NVM for packet buffers. DPDK uses a
concurrent memory allocator based on reference counting - used by both the NIC
driver and CPU cores. This means that packet allocation and deallocation does
not happen in the same order as their corresponding position in the replicated
RAFT log. To deal with this, we use a level of indirection as shown in
Figure~\ref{fig:nvm_log}. The NVM log is maintained as a circular log of fixed
sized pointers. Adding a level of indirection in the NVM log allows us to
separate the FIFO ordered circular log being manipulated by RAFT from packet
data being managed by the memory allocator of DPDK. Both the circular log and
packet data are in NVM. An advantage of this scheme is that it makes recovery
from NVM easy - appends to the circular pointer log are atomic and we can use
the pointer log to recover allocator state i.e. what pieces of NVM are currently
in use by the log.

@@

\subsection{Eliminating Data Movement}
Software packet switches built on DPDK try to touch as little of the packet data
as possible, minimizing movement of data up and down the cache and memory
hierarchy. Log replication looks like multicast that does not require deep
packet inspection. Software packet switches implementing multicast therefore
simply manipulate packet headers to produce new packets to send on the output
ports. We designed our implementation of raft around the same principle.

\begin{figure}
\begin{verbatim}
    event_receive_client_req()
    {
      if(!check_is_leader()) {
        drop_msg
        return
      }
      Prepend raft log term and index
      Persist to local log
      Transmit to follower replicas
    }
\end{verbatim}
\caption{Cyclone event handling}
\label{fig:control_plane}
\end{figure}

The pseudocode in Figure~\ref{fig:control_plane} describes part of the packet
handling code in Cyclone organized as event handlers triggered on receiving a
packet at the leader. We focus on only one key event for brevity: the event
where a request is received from a client. The first step is to check that this
replica is indeed the RAFT leader in the current view (term). If not, the
message is simply dropped (a timeout causes the client to try a different
server). It then prepends RAFT related information to the packet - this includes
the current term and log index for this entry. Next, it appends a pointer to the
log in Figure~\ref{fig:nvm_log} effectively appending the packet to the
persistent log. Finally, it transmits the packet to follower replicas.

This entire process is done without making any copies of the received packet.
To illustrate how this is achieved, Figure~\ref{fig:packet_layout} shows how
Cyclone manipulates packet
layouts across the two steps of prepending a RAFT header and transmitting to
follower replicas. DPDK describes packets using an ``mbuf'' data
structure. Roughly speaking, an ``mbuf'' consists of a flat array of bytes
actually containing the packet and a fixed size piece of external metadata that
describes various aspects of the packet, most crucially a pointer to the start
and end of the packet in the byte array. DPDK's userspace drivers receive
packets from the NIC such that they are offset in the byte array by a
configurable amount referred to as ``headroom''. We strip off the existing
network headers in the packet and prepend RAFT related information specific to
each log entry in the headroom by shifting the start pointer appropriately.
These operations are standard enough for software packet switches that DPDK
provides convenient library calls for it. For the final step, we need to prepare
the packet for transmission to the various follower replicas. To do this we
prepare a different packet containing an ethernet header for \emph{each}
targeted replica and ``chain'' the data packet to each of these headers. Each
header is then separately handed off to the driver for transmission via the NIC,
carrying the data packet with it by association.

@@

Finally, we turn our attention to the persistence step in
Figure~\ref{fig:control_plane}. RAFT requires that the log entry be persisted
before it is multicast out to follower replicas. DPDK userspace NIC drivers
operate in DDIO mode~\cite{ddio} where the packet is directly written into the
CPU cache rather than first being DMAed into DRAM and then fetched by the CPU on
demand. Since the NVM is directly attached we need to persist the packet to it
by executing a cacheline flush ({\tt clflush}) instruction for every cacheline
in the packet and the pointer in the pointer buffer to persist these via the
memory bus. This is not too onerous a burden because we can use the newly
introduced {\tt clflush-opt}~\cite{clflush_opt} instruction specifically
intended to efficiently flush to persistent memory without the overhead of the
serialization normally introduced by {\tt clflush}. This allows us to hit full
memory bandwidth on present generation platforms, a quantity in excess of 200
Gb/s per core, well above the near term speeds of network interface cards. We
execute a single serializing {\tt sfence} before the sequence of {\tt clflushes}
to make sure any dirty cacheline due to header manipulation related writes from
the CPU are sent to cache.

@@

Although RAFT is an efficient consensus protocol in the common case, the
protocol state machine still adds significant overhead to each packet, relative
to the time for the packet to flow in from the NIC and back out to the replicas.
We address this problem for the \emph{loaded} case using batching - treating a
whole sequence of client commands as a single RAFT log entry, while avoiding any
copies to group these packets together. Figure~\ref{fig:batching} illustrates
how this is done. We use a burst receive call available in the DPDK userspace
driver to receive a burst of client packets at a time. We then chain these
packets together and treat them as a single log entry from the perspective of
RAFT, amortizing the control plane overheads over the packets (at most 32 at a
time due to current driver limitations). Crucially, batching in Cyclone does not
involve a latency-throughput tradeoff like in many other
systems~\cite{ix-dataplane}. The batch receive call we use in DPDK returns
immediately with whatever number of packets is available, including zero. We
always flush the transmit buffer after every call to DPDK to transmit packets to
replicas. Therefore, we never tradeoff latency for throughput when batching.
